---
title:  "Species Sensitivity Distributions (SSD) for Ecotoxicological Risk Assessment"
author: "Hannes Reinwald"
date:   "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction
Species Sensitivity Distributions (SSD) are a widely used tool in ecotoxicology for assessing the risk of chemicals to various species. This document provides a short, self-contained example of the core SSD workflow using the [`ssdtools`](https://bcgov.github.io/ssdtools/) package: prepare data, fit distributions, plot the SSD, and derive a hazard concentration.

We use the built-in `ccme_boron` dataset so the example is fully reproducible. To apply this workflow to your own standartox-derived data, see the custom analysis in student task #04.

```{r}
# Load packages
library(ssdtools)
library(ssddata)
library(ggplot2)
```

## Data
SSD input is one aggregated toxicity value (e.g. a geometric-mean LC50/EC50) per species, in a single column. In `ccme_boron` that column is `Conc`.

```{r}
head(ccme_boron)
```

## Fit distributions
`ssd_fit_dists()` fits several candidate distributions at once (by default gamma, lgumbel, llogis, lnorm, lnorm_lnorm and weibull).

```{r}
fits <- ssd_fit_dists(ccme_boron)
fits
```

## Compare the fits
`ssd_gof()` reports goodness-of-fit statistics and model weights. `ssdtools` combines the candidate distributions via model averaging (weighted by these values) rather than relying on a single "best" fit.

```{r}
ssd_gof(fits, wt = TRUE)
```

## Plot the SSD
`ssd_plot_cdf()` draws the model-averaged cumulative distribution together with the observed species data points.

```{r}
ssd_plot_cdf(fits)
```

## Hazard concentration
`ssd_hc()` estimates hazard concentrations from the model-averaged fit. The HC5 is the concentration protective of 95% of species and is a common regulatory benchmark.

```{r}
ssd_hc(fits, proportion = 0.05)
```

## Next steps
For the regulatory BCANZ workflow (a fixed distribution set with bootstrap confidence intervals), use `ssd_fit_bcanz()` together with `ssd_hc_bcanz()`. To run an SSD on your own curated standartox dataset (`data/ssd_dataset.tsv`), follow student task #04.
